---
title: "Contingency Tables and Categorical Data"
author: Andreas Kitsche
output: github_document
---

When a sample of $n$ observations in two qualitative (categorical) variables are available, they can be arranged into a cross-classification in which the number of observations falling in each cell of the table is recorded. This table is also denoted as contingency table, in which the observations for a sample of individuals or objects are cross-classified whith respect to two categorical variables. Testing for the independence of the two variables $x$ and $y$ is of most interest in general.

Under the null hypothesis of independence of the row variable $x$ and the column variable $y$, estimated expected values $E_{jk}$ for cell $(j,k)$ can be computed from the corresponding margin totals $E_{jk}=n_{j.}n_{.k}/n_{.}$. The test statistic for assessing independence is

$$
\chi^{2}=\sum_{j=1}^{r}\sum_{k=1}^{c}\frac{(n_{jk}-E_{jk})^2}{E_{jk}}
$$
Under the null hypothesis of independence, the test statistic $\chi^{2}$ is asymptotically distributed according to a $\chi^{2}$-distribution with $(r-1)(c-1)$degrees of freedom, the corresponding test is usually known as chi-squared test. 

##Example - Goodness of fit test (One-variable $\chi^{2}$-test)
We think that the proportion of nucleotides A, T, C, G should be equally present in a given DNA sequence, with proportion 0.25 for each. Let's see for a particular gene how good this fit really is. For this example we will use the sequence for human myoglobin, a muscle protein which carries oxygen. The obtained number of each nucleotide is A-237, T-278, C-309, G-242. 
Formulate an appropriate null and alternative hypothesis:

$$
H_{0}:\hspace{0.5cm} \pi_{A}=0.25\hspace{0.5cm}  \pi_{G}=0.25\hspace{0.5cm} \pi_{C}=0.25\hspace{0.5cm} \pi_{T}=0.25
$$

$$
H_{A}:\hspace{0.5cm} H_{0} \hspace{0.5cm} is\hspace{0.5cm} false
$$

Plotting the frequency of each nucleotide using the `barplot()` function
```{r cache=TRUE, echo=TRUE, error=TRUE, fig.show='asis', message=TRUE, warning=TRUE, results='asis',tidy=TRUE}
obs <- c(237, 278, 309, 242)
barplot(obs, ylab="absolute frequencies of nucleotides", 
             main="Frequencies of nucleotides of Myoglobin", 
             names.arg=c("A","T","C","G"))
```

Calculating the corresponding chi square test with R
```{r cache=TRUE, echo=TRUE, error=TRUE, fig.show='asis', message=TRUE, warning=TRUE, results='asis',tidy=TRUE}
expected <- c(0.25, 0.25, 0.25, 0.25)
chisq.test(obs,p = expected)
```
```{r results='hide', message=FALSE, warning=FALSE, cache=TRUE, fig.show='asis', echo=FALSE,tidy=TRUE}
ChiTest <- chisq.test(obs,p = expected)
teststatistic <- ChiTest$statistic
teststatistic
#null hypothesis
x <-seq(-10,20, by=0.001)
curve(dchisq(x, df=3),from=-10, to=20, main="Density of the Chi square distribution with df=3", ylab="density")
#input teststatistic
lines(x=c(teststatistic, teststatistic), y=c(0,0.05))
text(x=10, y=0.08, "X-squared= 12.79174")
x.x <- seq(teststatistic, 20, by=0.001)
y.x <- dchisq(x.x, df=3)
polygon(x = c(teststatistic, x.x, 20, teststatistic), y=c(0, y.x, 0, 0), col="blue")
1-pchisq(q=teststatistic, df=3)
text(x=17, y=0.02, "p-value = 0.005")
```